Titeln på avhandlingens språk

### 2022-12-19 old_model.tex
glossaries

As expressed in \cref{ssec:plos}, for ~\cite{olguinmunoz:impact2021} we collected timing and task performance data for \num{40} participants performing a \num{169}-step task, guided by a \ac{WCA}, as well as personality trait scores for each participant.
For this paper, we use this data to construct a probabilistic model for the generation of realistic execution times.
The data is pre-processed to group execution time values by level of neuroticism, previous \acp{TTF}, and duration of impairment.
The resulting collections of execution times represent the distributions of these values for users with specific levels of neuroticism, interacting with systems at specific states of impairment and recent histories of impairment.
These distributions can then be sampled to produce new, realistic execution times.
In the following, we detail the step-by-step processing of this data and construction of the probabilistic model.

@@

We begin by discretizing normalized neuroticism into \emph{low} and \emph{high} ranges, and \ac{TTF} into three ranges by splitting the data on the 3-quantiles; we will refer to these ranges of \ac{TTF} as \emph{low}, \emph{medium}, and \emph{high impairment}.
\cref{tab:defaultbins} shows the exact values used for these ranges.

The next steps of processing are slightly more complex.
For the data of each subject (which we will refer to as a \emph{run}), we shift the execution times by one sample such that each \ac{TTF} measurement (and therefore, impairment level) is associated with the execution time for the next step.
This comes directly from our discussions in \cref{sec:background} and~\cite{olguinmunoz:impact2021}.
Users notice the impairment of the current step only when they have finished it, and thus system impairment has no effect on the execution time of the current step, but directly affects the next one.

@@

We use Python~\num{3.10} to implement two variants of the above described timing model.
The first of these samples execution times directly from the empirical data from~\cite{olguinmunoz:impact2021}; we will refer to it hereafter as the \emph{empirical model}.
The second corresponds to one which first fits a theoretical \ac{exGaussian} distribution to the data before sampling; we will refer to this one as the \emph{theoretical model}.
The \ac{exGaussian} distribution was chosen due to the vast body of research supporting its suitability for the modeling of the timing of human actions~\cite{Rohrer1994analysis,Palmer2011shapes,Marmolejo2022generalised}.
Collectively, we will refer to both these models interchangeably as the \emph{realistic models} or \emph{second-order models}.
\todo[inline]{Explain why using ExGaussian?}

We also implement two simpler reference timing models:
\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
    % \item a model which returns a constant execution time, corresponding to the mean execution time of all samples associated with the lowest impairment and the shortest duration
    \item a model which randomly samples from the raw data without any grouping
    \item a model which fits an \ac{exGaussian} distribution to the raw data, without any grouping, and then randomly samples from it  
\end{enumerate*}.
We will refer to these models individually as the \emph{naive} and \emph{fitted-naive} model, respectively, and collectively as the \emph{firt-order models}.

\begin{figure*}
    \centering
    \includegraphics[width=.9\textwidth]{figs/model_comp_controlled_ttf.png}
    \begin{tabular}{ll >{\columncolor[HTML]{EAEAEA}}r >{\columncolor[HTML]{EAEAEA}}r rr}
    \toprule
    {} & {} & \multicolumn{2}{>{\columncolor[HTML]{EAEAEA}}c}{\ac{TTF} \( = \) \SI{0}{\second}} & \multicolumn{2}{c}{\ac{TTF} \( = \) \SI{5}{\second}} \\
    {Model} & {Neuroticism} & {Total Duration} & {Diff.\ w.r.t.\ fitted-naive} & {Total Duration} & {Diff.\ w.r.t.\ fitted-naive} \\
    \midrule
    fitted-naive & & \SI{575.54}{\second} & & \SI{1082.28}{\second} & \\
    \multirow[c]{2}{*}{empirical} & low & \SI{509.39}{\second} & \SI{-11.49}{\percent} & \SI{1071.99}{\second} & \SI{-0.95}{\percent} \\
    & high & \SI{536.48}{\second} & \SI{-6.79}{\percent} & \SI{1235.65}{\second} & +\SI{14.17}{\percent} \\
    \multirow[c]{2}{*}{theoretical} & low & \SI{505.22}{\second} & \SI{-12.22}{\percent} & \SI{1073.49}{\second} & \SI{-0.81}{\percent} \\
    & high & \SI{536.31}{\second} & \SI{-6.82}{\percent} & \SI{1235.70}{\second} & +\SI{14.18}{\percent} \\
    \bottomrule
    \end{tabular}
    \caption{
        Comparison of the total duration of \num{100}-step tasks for different models in the controlled \ac{TTF} setup.
        The table shows a detailed breakdown of durations for the fitted-naive, empirical, and theoretical models at the lowest and highest \acp{TTF}, as well as relative differences with respect to the corresponding fitted-naive baseline for each case.
    }\label{fig:ctrlttf}
\end{figure*}

We first compare these models using a controlled, ideal setup with \num{25} constant, linearly distributed \acp{TTF} in the \SIrange[]{0}{5}{\second} range.
For each \ac{TTF}, the models execute \num{30} independent repetitions of a \num{100} step task.

We show the results of these runs in \cref{fig:ctrlttf}.
Two interesting behaviors of the realistic models can be noticed in this plot.
One, all realistic models are faster than the reference models at the lowest \ac{TTF}.
When compared to the fitted-naive model, the reduction in total duration was on average \textasciitilde\SI{12}{\percent} for realistic models with low neuroticism, and \textasciitilde\SI{6.8}{\percent} for the same models with high neuroticism.

Two, at the highest \ac{TTF}, \emph{only} the high neuroticism models show any significant difference in duration with respect to the reference models.
For \ac{TTF} \(=\) \SI{5.0}{s}, the high neuroticism realistic models were on average \textasciitilde\SI{14.18}{\percent} slower than the fitted-naive model, whereas the low neuroticism models showed no significant relative difference in duration.

@@

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figs/EdgeDroid2ExperimentalSetup.png}
    \caption{%
        Experimental setup used to study the implications of the realistic models of human behavior for \ac{WCA}.
        We deploy containerized instances of the client-server loop running the models on a testbed consisting of \num{10} Raspberry Pi clients connected to a cloudlet over a \ac{COTS} \acs{IEEE} \num{802.11}n access point.
    }\label{fig:expsetup}
\end{figure}

These models are then integrated into client software for communicating with an adapted and instrumented version of the Gabriel~\cite{Chen2015LEGO,Chen2018application} \ac{WCA} backend running variants of the LEGO task.
In the following we use these client-server loops together with the testbed show in \cref{fig:expsetup} to illustrate the implications of the realistic timing models.
Our goal is to understand the consequences of taking second order effects into consideration when modeling execution times, versus simply relying on first-order statistics.

@@

\begin{figure*}
    \centering
    \begin{subfigure}[t]{\columnwidth}
        \centering
        \includegraphics[width=.8\textwidth]{figs/numclients_vs_duration.png}
        \caption{%
            Consequences of decreased system responsiveness due to resource and network congestion.
            At lower levels of resource contention, the realistic models outperform the naive one due to the speed-up effect described in \cref{sec:background}.
            However, at higher levels of contention, the behavioral slow-down of the realistic models due to higher \acp{TTF} takes over.
        }\label{fig:scaling_duration}
    \end{subfigure}%
    \hspace{\fill}%
    \begin{subfigure}[t]{\columnwidth}
        \centering
        \includegraphics[width=.8\textwidth]{figs/neuroticism_vs_duration.png}
        \caption{%
            Modulating effects of neuroticism on task duration.
            Realistic models parameterized with high neuroticism are on average \SI{7.3}{\percent} slower than the reference model, and \SI{3.37}{\percent} slower than the same models instead parameterized with low neuroticism.
        }\label{fig:neuro_duration}
    \end{subfigure}%
    \caption{Results for both scenarios described in \cref{ssec:model:verification}. Error bars indicate the \SI{95}{\percent} \ac{CI}.}
\end{figure*}

@@

The results for scenarios \labelcref{verification:scaling,verification:neuro} showcase the importance and potential impact of realistic models for the study of \ac{WCA} applications.

>>>>

As expressed in \cref{ssec:plos}, for ~\cite{olguinmunoz:impact2021} we collected timing and task performance data for \num{40} participants performing a \num{169}-step task, guided by a \gls{WCA}, as well as personality trait scores for each participant.
For this paper, we use this data to construct a probabilistic model for the generation of realistic execution times.
The data is pre-processed to group execution time values by level of neuroticism, previous \acp{TTF}, and duration of impairment.
The resulting collections of execution times represent the distributions of these values for users with specific levels of neuroticism, interacting with systems at specific states of impairment and recent histories of impairment.
These distributions can then be sampled to produce new, realistic execution times.
In the following, we detail the step-by-step processing of this data and construction of the probabilistic model.

@@

We begin by discretizing normalized neuroticism into \emph{low} and \emph{high} ranges, and \gls{TTF} into three ranges by splitting the data on the 3-quantiles; we will refer to these ranges of \gls{TTF} as \emph{low}, \emph{medium}, and \emph{high impairment}.
\cref{tab:defaultbins} shows the exact values used for these ranges.

The next steps of processing are slightly more complex.
For the data of each subject (which we will refer to as a \emph{run}), we shift the execution times by one sample such that each \gls{TTF} measurement (and therefore, impairment level) is associated with the execution time for the next step.
This comes directly from our discussions in \cref{sec:background} and~\cite{olguinmunoz:impact2021}.
Users notice the impairment of the current step only when they have finished it, and thus system impairment has no effect on the execution time of the current step, but directly affects the next one.

@@

We use Python~\num{3.10} to implement two variants of the above described timing model.
The first of these samples execution times directly from the empirical data from~\cite{olguinmunoz:impact2021}; we will refer to it hereafter as the \emph{empirical model}.
The second corresponds to one which first fits a theoretical \gls{exGaussian} distribution to the data before sampling; we will refer to this one as the \emph{theoretical model}.
The \gls{exGaussian} distribution was chosen due to the vast body of research supporting its suitability for the modeling of the timing of human actions~\cite{Rohrer1994analysis,Palmer2011shapes,Marmolejo2022generalised}.
Collectively, we will refer to both these models interchangeably as the \emph{realistic models} or \emph{second-order models}.
\todo[inline]{Explain why using ExGaussian?}

We also implement two simpler reference timing models:
\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
    % \item a model which returns a constant execution time, corresponding to the mean execution time of all samples associated with the lowest impairment and the shortest duration
    \item a model which randomly samples from the raw data without any grouping
    \item a model which fits an \gls{exGaussian} distribution to the raw data, without any grouping, and then randomly samples from it
\end{enumerate*}.
We will refer to these models individually as the \emph{naive} and \emph{fitted-naive} model, respectively, and collectively as the \emph{firt-order models}.

\begin{figure*}
    \centering
    \includegraphics[width=.9\textwidth]{figs/model_comp_controlled_ttf.png}
    \begin{tabular}{ll >{\columncolor[HTML]{EAEAEA}}r >{\columncolor[HTML]{EAEAEA}}r rr}
    \toprule
    {} & {} & \multicolumn{2}{>{\columncolor[HTML]{EAEAEA}}c}{\gls{TTF} \( = \) \SI{0}{\second}} & \multicolumn{2}{c}{\gls{TTF} \( = \) \SI{5}{\second}} \\
    {Model} & {Neuroticism} & {Total Duration} & {Diff.\ w.r.t.\ fitted-naive} & {Total Duration} & {Diff.\ w.r.t.\ fitted-naive} \\
    \midrule
    fitted-naive & & \SI{575.54}{\second} & & \SI{1082.28}{\second} & \\
    \multirow[c]{2}{*}{empirical} & low & \SI{509.39}{\second} & \SI{-11.49}{\percent} & \SI{1071.99}{\second} & \SI{-0.95}{\percent} \\
    & high & \SI{536.48}{\second} & \SI{-6.79}{\percent} & \SI{1235.65}{\second} & +\SI{14.17}{\percent} \\
    \multirow[c]{2}{*}{theoretical} & low & \SI{505.22}{\second} & \SI{-12.22}{\percent} & \SI{1073.49}{\second} & \SI{-0.81}{\percent} \\
    & high & \SI{536.31}{\second} & \SI{-6.82}{\percent} & \SI{1235.70}{\second} & +\SI{14.18}{\percent} \\
    \bottomrule
    \end{tabular}
    \caption{
        Comparison of the total duration of \num{100}-step tasks for different models in the controlled \gls{TTF} setup.
        The table shows a detailed breakdown of durations for the fitted-naive, empirical, and theoretical models at the lowest and highest \acp{TTF}, as well as relative differences with respect to the corresponding fitted-naive baseline for each case.
    }\label{fig:ctrlttf}
\end{figure*}

We first compare these models using a controlled, ideal setup with \num{25} constant, linearly distributed \acp{TTF} in the \SIrange[]{0}{5}{\second} range.
For each \gls{TTF}, the models execute \num{30} independent repetitions of a \num{100} step task.

We show the results of these runs in \cref{fig:ctrlttf}.
Two interesting behaviors of the realistic models can be noticed in this plot.
One, all realistic models are faster than the reference models at the lowest \gls{TTF}.
When compared to the fitted-naive model, the reduction in total duration was on average \textasciitilde\SI{12}{\percent} for realistic models with low neuroticism, and \textasciitilde\SI{6.8}{\percent} for the same models with high neuroticism.

Two, at the highest \gls{TTF}, \emph{only} the high neuroticism models show any significant difference in duration with respect to the reference models.
For \gls{TTF} \(=\) \SI{5.0}{s}, the high neuroticism realistic models were on average \textasciitilde\SI{14.18}{\percent} slower than the fitted-naive model, whereas the low neuroticism models showed no significant relative difference in duration.

@@

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figs/EdgeDroid2ExperimentalSetup.png}
    \caption{%
        Experimental setup used to study the implications of the realistic models of human behavior for \gls{WCA}.
        We deploy containerized instances of the client-server loop running the models on a testbed consisting of \num{10} Raspberry Pi clients connected to a cloudlet over a \gls{COTS} \acs{IEEE} \num{802.11}n access point.
    }\label{fig:expsetup}
\end{figure}

These models are then integrated into client software for communicating with an adapted and instrumented version of the Gabriel~\cite{Chen2015LEGO,Chen2018application} \gls{WCA} backend running variants of the LEGO task.
In the following we use these client-server loops together with the testbed show in \cref{fig:expsetup} to illustrate the implications of the realistic timing models.
Our goal is to understand the consequences of taking second order effects into consideration when modeling execution times, versus simply relying on first-order statistics.

@@

\begin{figure*}
    \centering
    \begin{subfigure}[t]{\columnwidth}
        \centering
        \includegraphics[width=.8\textwidth]{figs/numclients_vs_duration.png}
        \caption{%
            Consequences of decreased system responsiveness due to resource and network congestion.
            At lower levels of resource contention, the realistic models outperform the naive one due to the speed-up effect described in \cref{sec:background}.
            However, at higher levels of contention, the behavioral slow-down of the realistic models due to higher \acp{TTF} takes over.
        }\label{fig:scaling_duration}
    \end{subfigure}%
    \hspace{\fill}%
    \begin{subfigure}[t]{\columnwidth}
        \centering
        \includegraphics[width=.8\textwidth]{figs/neuroticism_vs_duration.png}
        \caption{%
            Modulating effects of neuroticism on task duration.
            Realistic models parameterized with high neuroticism are on average \SI{7.3}{\percent} slower than the reference model, and \SI{3.37}{\percent} slower than the same models instead parameterized with low neuroticism.
        }\label{fig:neuro_duration}
    \end{subfigure}%
    \caption{Results for both scenarios described in \cref{ssec:model:verification}. Error bars indicate the \SI{95}{\percent} \gls{CI}.}
\end{figure*}

@@

The results for scenarios \labelcref{verification:scaling,verification:neuro} showcase the importance and potential impact of realistic models for the study of \gls{WCA} applications.

### 2022-12-19 related_work.tex
vishnu
glossaries

There exist a number of approaches to the characterization and benchmarking of mobile \ac{XR}, and by extension \ac{WCA}, deployments.
\emph{OpenRTiST}~\cite{george2020openrtist} is a tool which uses a compute-heavy, yet latency-sensitive workload --- image style transfer~\cite{jing2019neural} --- to load and benchmark edge computing deployments.
It acts as an end-to-end realistic \ac{AR} workload, allowing thus researchers to measure the real end-to-end latencies of such systems.
\textcite{Chen2017Empirical} use a number of prototype \ac{WCA} implementation to study and characterize real-world latency bounds in these systems.
In~\cite{lecci2021open}, the authors collect and analyze large amounts of \ac{VR} network traffic, which they then use to construct a synthetic model for the generation of traces of such traffic.
\textcite{chetoui2022arbench} propose \emph{ARBench}, a toolkit for for the benchmarking of mobile hardware in the context of \ac{AR}.
The toolkit incorporates a series of workloads which stress different components of the mobile device and calculates a score for each workload.
\emph{Yahoo Cloud Serving Benchmark}~\cite{cooper2010benchmarking} and \emph{DCBench}~\cite{jia2013characterizing} are workload sets intended for benchmarking cloud services which have further been used in the context of benchmarking edge computing infrastructure for \ac{AR} suitability.
\emph{Edgebench}~\cite{das2018edgebench} and \emph{Defog}~\cite{mcchesney2019defog} benchmark workload performance on the edge versus on the cloud.

We contribute to this body of work by providing the first tool for the benchmarking of \ac{WCA} that dynamically considers the effects of human behavior.
In contrast to the above described works, which generate loads with relatively static profiles, our model is able to react to changes in system responsiveness like a human would.
We achieve this by building upon our own previous contributions to this field.
In~\cite{olguin2018scaling,olguin2019edgedroid} we introduced a coarse approximation to human behavior modeling in \ac{WCA} we called \emph{EdgeDroid}.
We used a trace-driven approach, where a pre-recorded and pre-processed ``ideal'' trace of steps for a specific \ac{WCA} step-based task is replayed to a Gabriel~\cite{Chen2018application} backend.
In order to adapt to potential mismatches between system responsiveness at trace-capture time and trace-replay time, we used a simple \ac{FSM} to adapt the trace at the latter by replaying or skipping certain segments.
In~\cite{olguinmunoz:impact2021} we studied the effects of reduced system responsiveness on human behavior in \ac{WCA} applications through human-subject studies.
We found that humans generally pace themselves according to the perceived system responsiveness.
When interacting with a highly responsive system, humans tended to speed up with each step; conversely, humans tended to slow down in highly unresponsive states.

The characterization and optimization of mobile \ac{XR} has also long been a topic of research.
\textcite{srinivasan2009performance} analyze a \ac{MAR} workload and identify and characterize the computational bottlenecks in the application.
They use this characterization to develop a series of code optimizations which allow them to achieve a threefold increase in performance.
\textcite{Wang2019Towards} develop a taxonomy for \ac{WCA} as well as strategies for workload reduction in these applications, including a novel adaptive sampling scheme with the goal of reducing the number of processed samples while maintaining application responsiveness.
\textcite{huang2021proactive} present a framework for the proactive allocation of edge cloud resources while taking into account the inherent mobility of wearable and mobile \ac{AR}.
\textcite{al2017energy} propose a resource allocation strategy which leverages collaboration between \ac{AR} applications to minimize energy consumption.
As above, we have also previously directly contributed to this field.
In~\cite{Moothedath2021EnergyOptimal,Moothedath2022EnergyEfficient} we find the optimum periodic sampling interval which minimizes the energy tracking human progress of a specific subtask in \ac{WCA}.
This is extended into an aperiodic strategy in~\cite{Moothedath2022Aperiodic}.
We further develop this approach in this work by deriving a more practical approximate solution to the aperiodic strategy, which we then use to implement a novel framework for the generic optimization of responsiveness-resource consumption trade-offs.

>>>>

There exist a number of approaches to the characterization and benchmarking of mobile \gls{XR}, and by extension \gls{WCA}, deployments.
\emph{OpenRTiST}~\cite{george2020openrtist} is a tool which uses a compute-heavy, yet latency-sensitive workload --- image style transfer~\cite{jing2019neural} --- to load and benchmark edge computing deployments.
It acts as an end-to-end realistic \gls{AR} workload, allowing thus researchers to measure the real end-to-end latencies of such systems.
\textcite{Chen2017Empirical} use a number of prototype \gls{WCA} implementation to study and characterize real-world latency bounds in these systems.
In~\cite{lecci2021open}, the authors collect and analyze large amounts of \gls{VR} network traffic, which they then use to construct a synthetic model for the generation of traces of such traffic.
\textcite{chetoui2022arbench} propose \emph{ARBench}, a toolkit for the benchmarking of mobile hardware in the context of \gls{AR}.
The toolkit incorporates a series of workloads which stress different components of the mobile device and calculates a score for each workload.
\emph{Yahoo Cloud Serving Benchmark}~\cite{cooper2010benchmarking} and \emph{DCBench}~\cite{jia2013characterizing} are workload sets intended for benchmarking cloud services which have further been used in the context of benchmarking edge computing infrastructure for \gls{AR} suitability.
\emph{Edgebench}~\cite{das2018edgebench} and \emph{Defog}~\cite{mcchesney2019defog} benchmark workload performance on the edge versus on the cloud.

We contribute to this body of work by providing the first tool for the benchmarking of \gls{WCA} that dynamically considers the effects of human behavior.
In contrast to the above described works, which generate loads with relatively static profiles, our model is able to react to changes in system responsiveness like a human would.
We achieve this by building upon our own previous contributions to this field.
In~\cite{olguin2018scaling,olguin2019edgedroid} we introduced a coarse approximation to human behavior modeling in \gls{WCA} we called \emph{EdgeDroid}.
We used a trace-driven approach, where a pre-recorded and pre-processed ``ideal'' trace of steps for a specific \gls{WCA} step-based task is replayed to a Gabriel~\cite{Chen2018application} backend.
In order to adapt to potential mismatches between system responsiveness at trace-capture time and trace-replay time, we used a simple \gls{FSM} to adapt the trace at the latter by replaying or skipping certain segments.
In~\cite{olguinmunoz:impact2021} we studied the effects of reduced system responsiveness on human behavior in \gls{WCA} applications through human-subject studies.
We found that humans generally pace themselves according to the perceived system responsiveness.
When interacting with a highly responsive system, humans tended to speed up with each step; conversely, humans tended to slow down in highly unresponsive states.

The characterization and optimization of mobile \gls{XR} has also long been a topic of research.
\textcite{srinivasan2009performance} analyze a \gls{MAR} workload and identify and characterize the computational bottlenecks in the application.
They use this characterization to develop a series of code optimizations which allow them to achieve a threefold increase in performance.
\textcite{Wang2019Towards} develop a taxonomy for \gls{WCA} as well as strategies for workload reduction in these applications, including a novel adaptive sampling scheme with the goal of reducing the number of processed samples while maintaining application responsiveness.
\textcite{huang2021proactive} present a framework for the proactive allocation of edge cloud resources while taking into account the inherent mobility of wearable and mobile \gls{AR}.
\textcite{al2017energy} propose a resource allocation strategy which leverages collaboration between \gls{AR} applications to minimize energy consumption.
As above, we have also previously directly contributed to this field.
In~\cite{Moothedath2021EnergyOptimal,Moothedath2022EnergyEfficient} we find the optimum periodic sampling interval which minimizes the energy tracking human progress of a specific subtask in \gls{WCA}.
This is extended into an aperiodic strategy in~\cite{Moothedath2022Aperiodic}.
We further develop this approach in this work by deriving a more practical approximate solution to the aperiodic strategy, which we then use to implement a novel framework for the generic optimization of responsiveness-resource consumption trade-offs.
